A hyperdimensional sequence code inspired by chromosomal crossover

### 2019-09-01 ms/crossing_over.tex
Rename.

\begin{abstract}
I present a method for encoding structured information in fixed-length vectors. Drawing on ideas from high-dimensional (HD) computing, the presented scheme encodes elements of information as HD random vectors, and relationships among them (like sequences or trees) by (pseudo-)randomly weaving these vectors together via an operation analogous to chromosomal crossover. Unlike previous HD coding schemes, "crossover" codes retain low Hamming distance to the multi-level sub-relationships that comprise them--all the way down to their base elements--yet high distance to alternative elements or relationships. Accurate decoding can thus occur via greedy reconstruction of the representation. Unlike most machine-learning approaches, crossover (as in other HD computing schemes) requires no training, making it highly amenable to mathematical analysis. I derive several of its formal properties and discuss its relevance to neuroscience and artificial intelligence.
\end{abstract}

% keywords can be removed
\keywords{Hyperdimensional computing \and Vector symbolic architectures \and Sequence encoding \and k-ary tree encoding \and Computational neuroscience}

\section{Introduction}
A core problem in neuroscience is how the brain represents a complex event like a spoken sentence using relatively fixed dimensions. Ideally similar events should have similar representations, and when a new event occurs pre-existing representations shouldn't require substantial rearrangement to prevent confusion.

@@

\subsection{Sequence codes have i.i.d. components}
\label{sec:iid-comp}

\textbf{Result:}

@@

??? A corollary is that if $Y \sim P(Y)$ then $P(x^Y_j) = P(x^Y_{j'}) \quad \forall j, j' \in [1, ..., N]^2$, and similarly $P(X^Y) $

\textbf{Proof:}

@@

\subsection{Component distributions are uniform}
\label{sec:prop-eqty-comp-dstr}

\textbf{Result:}

\textbf{Proof:}

@@

\textbf{Result:}

\textbf{Proof:}

This follows from \ref{sec:prop-eqty-comp-dstr}, since $P(x^Y_j|Y) = 1/Z$. Therefore the statistics of crossover codes do not depend on any aspect about the sequence being encoded. This distinguishes crossover, for instance, from representations that grow more or less crowded with the complexity of the sequence they represent.

\subsection{Information about sequence identity is evenly distributed across the code}
\label{sec:perf-info-dstr}

Given $B$, $\mu$, and a distribution over sequences $P_Y(Y)$ with entropy $H[Y]$, define the information $(I_j | B, \mu)$ in component $j$ as the mutual information between $Y$ and $x_j$, i.e.

$$I_j | B, \mu \equiv MI[Y|x_j; B, \mu] = H[Y] - H[Y|x_j; B, \mu] = H[Y] - \sum_g H[Y|x_j = g; B, \mu] P_Y(x_j = g| B, \mu)$$

\textbf{Result:}

\textbf{Proof:}

This follows from \ref{sec:iid-comp}, since the components for one sequence $Y$ are i.i.d. Therefore, while the $x^Y_j$ and $x^{Y'}_j$ might be dependent (e.g. if $Y$ and $Y'$ are very similar; see below), there can be no dependencies between $x^Y_j$ and $x^{Y'}_{j'}$ for $j \neq j'$. Therefore $\{x^{Y_1}_j, ..., x^{Y_{N_Y}}_j\}$ is independent across $j$. Since $I_j$ is a function of $\{x^{Y_1}_j, ..., x^{Y_{N_Y}}_j\}$, $I_j$ is independent across $j$ also. Since no $j$ is treated specially in our algorithm, $I_j$ must also be identically distributed, so given $P(Y)$, $I_j$ is i.i.d.

\subsection{Similarity between sequences}

\textbf{Result:}

Since $x^Y_j$ are i.i.d. across $j$ (\ref{sec:iid-comp}) it follows that $s(X^Y, X^{Y'}) \sim \frac{1}{N}\textrm{Bi}(N, q)$, where $\textrm{Bi}(N, q)$ is a binomial distribution with success probability $q = P(x^Y_j = x^{Y'}_j)$. Thus, $\textrm{E}[s(X^Y, X^{Y'})] = q$ and $\textrm{Var}[s(X^Y, X^{Y'})] = q(1-q)/N$, and it suffices to only compute $q$. For conciseness, we drop the $j$ and write $x \equiv x_j$ from now on.

\textbf{Proof:}

\textbf{Prelude:}

We first define and compute a few useful quantities that will simplify the rest of our calculations.

@@

\textbf{Similarity between symbol and sequence codes:}

If $y_t = i$ for $0 < t_1 < ... < t_{n^i}$, then

$$P(x^Y = x^i) = \frac{1}{Z} + \sum\limits_{t \in (t_1, ..., t_{n^i})}\left(1 - \frac{1}{Z}\right)f^{\gamma,L}_t.$$

For $\gamma = 1$:

$$P(x^Y = x^i) = \frac{n}{L+1} + \frac{1}{Z}\frac{L+1-n}{L+1}$$

which tends to $n/(L+1)$ for large $Z$.

\textbf{Proof}:

$$P(x^Y = x^i) = P(x^Y \leftarrow i)P(x^Y = x^i|x^Y \leftarrow i) + P(x^Y \not \leftarrow i)P(x^Y = x^i|x^Y \not \leftarrow i)$$

$$= P(x^Y \leftarrow i) + P(x^Y \not \leftarrow i)\frac{1}{Z}$$

@@

\textbf{Similarity between codes for two sequences:}

Let $Y = (y_1, ..., y_L)$ and $Y' = (y'_1, .., y'_{L'})$ be two sequences. Suppose they share a starting subsequence of length $t$, i.e. $y_1 = y'_1, ..., y_t = y'_t$, but either $y_{t+1} \neq y'_{t+1}$, $t = L$, or $t = L'$. We write this condition as $Y_{1:t} = Y'_{1:t}$. Similarly, we refer to the second parts of the sequences (the subsequences in which they differ) as $Y_{t+1:L}$ and $Y'_{t+1:L'}$, respectively.

Let $u^i$ index the times $i$ appears in $Y_{1:t}$, $v^i$ the times $i$ appears in $Y_{t+1:L}$, and $(v')^i$ the times $i$ appears in $Y'_{t+1:L'}$. Let $n^i_u$ count how many times $i$ appears in $Y_{1:t}$, $n^i_v$ the number of times $i$ appears in $Y_{t+1:L}$, and $n^i_{v'}$ the number of times $i$ appears in $Y'_{t+1:L'}$. We also write these in vector format as $\mathbf{n}_u$, $\mathbf{n}_v$, and $\mathbf{n}_{v'}$, respectively, where each is a vector of length $M$.

Note that we must deal specially with starting subsequences because sequence codes are constructed from beginning to end.

Then we have:

Wrong: Note that when $Y$ and $Y'$ do not start with the same subsequence (i.e. $t = 0$) this reduces to

$$P(x^Y = x^{Y'}) = \frac{1}{(L+1)(L'+1)}\left[1 + \frac{L + L' + LL'}{Z} + \frac{Z-1}{Z}\mathbf{n}^T\mathbf{n}'\right]$$

\textbf{Proof:}

@@

The event $x^Y = x^{Y'}$ can occur in 4 distinct ways, whose probabilities sum to $P(x^Y = x^{Y'})$. At present we work under the assumption that all symbol codes are different and that $Y \neq Y' \implies X^{Y} \neq X^{Y'}.$

@@

$$= [1 - (t+1)^\gamma f^{\gamma,L}_t][1 - (t+1)^\gamma f^{\gamma,L'}_t]P(x^Y = x^{Y'}|x^Y, x^{Y'} \not\leftarrow Y_{1:t})

@@

\textbf{Sum over all cases}

@@

When $\gamma = 1$, then $P(x^Y = x^{Y'}) = $

$$\frac{t+1}{L+1}\frac{t+1}{L'+1}
+ \frac{L-t}{L+1}\frac{L'-t}{L'+1}\left[\frac{\mathbf{n}^T_v\mathbf{n}_{v'}}{(L-t)(L'-t)} + \left(1 - \frac{\mathbf{n}^T_v\mathbf{n}_{v'}}{(L-t)(L'-t)} \right)\frac{1}{Z}\right]
$$

$$+\frac{t+1}{L+1}\frac{L'-t}{L'+1}\left[
\frac{\mathbf{n}^T_{u}\mathbf{n}_{v'}}{(t+1)(L'-t)} + \left(1 - \frac{\mathbf{n}^T_{u}\mathbf{n}_{v'}}{(t+1)(L'-t)} \right)\frac{1}{Z}
\right]
$$

$$+\frac{L'-t}{L+1}\frac{t+1}{L'+1}\left[
\frac{\mathbf{n}^T_{v}\mathbf{n}_{u}}{(L-t)(t+1)} + \left(1 - \frac{\mathbf{n}^T_{v}\mathbf{n}_{u}}{(L-t)(t+1)} \right)\frac{1}{Z}
\right]
$$

\textbf{OLD STARTS HERE}

This means that when $Y'$ is a starting subsequence of $Y$, i.e. $Y = (y'_1, ..., y'_t, y_{t+1}, ..., y_L)$ then the two sequences' expected similarity is equal to

Blah blah

>>>>

\begin{abstract}
I present a method for encoding structured information in fixed-length vectors. Drawing on ideas from high-dimensional (HD) computing, the presented scheme encodes elements of information as HD random vectors and sequences of elements by (pseudo-)randomly weaving these vectors together via an operation analogous to chromosomal crossover. "Crossover" codes retain low Hamming distance to both their elements and the codes for the subsequences used to build them, yet high distance to alternative elements or sequences. Accurate decoding can thus occur via greedy reconstruction of the representation. Unlike most machine-learning approaches, crossover (as in other HD computing schemes) requires no training, making it highly amenable to mathematical analysis. I derive several properties of crossover and discuss its relevance to neuroscience and artificial intelligence.
\end{abstract}

% keywords can be removed
\keywords{Hyperdimensional computing \and Vector symbolic architectures \and Sequence encoding \and Computational neuroscience \and Artificial intelligence \and Cognitive computing}

\section{Introduction}
A core problem in neuroscience and artificial intelligence is how to represent a complex event like a spoken sentence using relatively fixed dimensions. Ideally similar events should have similar representations, and when a new event occurs pre-existing representations shouldn't require substantial rearrangement to prevent confusion.

@@

\subsection{Sequence codes have i.i.d. components}
\label{sec:prop-comp-iid}

@@

\subsection{Component distributions are uniform over c-states}
\label{sec:prop-unif-comp-dstr}

\subsection{Sequence information is i.i.d. across components}
\label{sec:prop-info-iid}

Given $B$, $\mu$, and a distribution over sequences $P_Y(Y)$ with entropy $H[Y]$, define the information $I_j | B, \mu$ in component $j$ as the mutual information between $Y$ and $x_j$, i.e.

$$I_j | B, \mu \equiv MI[Y|x_j; B, \mu] = H[Y] - H[Y|x_j; B, \mu] = H[Y] - \sum_g H[Y|x_j = g; B, \mu] P_Y(x_j = g| B, \mu).$$

Then:

\section{Code similarities}

We give results for $\gamma = 1$ here, with results for arbitrary $\gamma$ in the appendix.

Since $x^Y_j$ are i.i.d. across $j$ (\ref{sec:prop-comp-iid}) it follows that $s(X^Y, X^{Y'}) \sim \frac{1}{N}\textrm{Bi}(N, q)$, where $\textrm{Bi}(N, q)$ is a binomial distribution with success probability $q = P(x^Y_j = x^{Y'}_j)$. Thus, $\textrm{E}[s(X^Y, X^{Y'})] = q$ and $\textrm{Var}[s(X^Y, X^{Y'})] = q(1-q)/N$, and it suffices to only compute $q$. For conciseness, we drop the $j$ and write $x \equiv x_j$ from now on.

\subsection{Similarity between symbol and sequence codes}

If there are $n^i$ repetitions of symbol $i$ appears in $Y$, then:

$$P(x^Y = x^i) = \frac{n^i}{L+1} + \frac{1}{Z}\frac{L+1-n^i}{L+1}$$

which tends to $n^i/(L+1)$ for large $Z$.

\subsection{Similarity between codes for two sequences}

Let $Y = (y_1, ..., y_L)$ and $Y' = (y'_1, .., y'_{L'})$ be two sequences that share a starting subsequence of length $t$. I.e. $y_1 = y'_1, ..., y_t = y'_t$, but either $y_{t+1} \neq y'_{t+1}$, $t = L$, or $t = L'$. We write that $Y_{1:t} = Y'_{1:t}$, and we refer to the second parts of the sequences as $Y_{t+1:L}$ and $Y'_{t+1:L'}$, respectively. We must deal specially with starting subsequences because sequence codes are constructed from beginning to end.

Let $\mathbf{n}_u$ be the length-$M$ symbol count vector for $Y_{1:t} = Y'_{1:t}$ with $\mathbf{1}^T\mathbf{n}_u = t$. Let $\mathbf{n}_v$ and $\mathbf{n}_{v'}$ be the symbol count vectors for $Y_{t+1:L}$ and $Y'_{t+1:L'}$, respectively, with $\mathbf{1}^T\mathbf{n}_{v} = L - t$ and $\mathbf{1}^T\mathbf{n}_{v'} = L' - t$. Then:

$$P(x^Y = x^{Y'}) = \frac{t+1}{L+1}\frac{t+1}{L'+1}$$

$$+ \frac{1}{L+1}\frac{1}{L'+1}\left[\mathbf{n}^T_v\mathbf{n}_{v'} + \left((L-t)(L'-t) - \mathbf{n}^T_v\mathbf{n}_{v'} \right)\frac{1}{Z}\right]
$$

$$+\frac{1}{L+1}\frac{1}{L'+1}\left[
\mathbf{n}^T_{u}\mathbf{n}_{v'} + \left((t+1)(L'-t) - \mathbf{n}^T_{u}\mathbf{n}_{v'} \right)\frac{1}{Z}
\right]
$$

$$+\frac{1}{L+1}\frac{1}{L'+1}\left[
\mathbf{n}^T_{v}\mathbf{n}_{u} + \left((L-t)(t+1) - \mathbf{n}^T_{v}\mathbf{n}_{u} \right)\frac{1}{Z}
\right]
$$

This means that when $Y'$ is a starting subsequence of $Y$, i.e. $Y' = Y_{1:t}$ then the two sequences' expected similarity is

\subsection{Proof: Sequence codes have i.i.d. components}

@@

\subsection{Proof: Component distributions are uniform over c-states}

@@

\subsection{Proof: Code statistics are invariant to sequence statistics}

This follows immediately from \ref{sec:prop-unif-comp-dstr}, since $P(x^Y_j|Y) = 1/Z$. Therefore the statistics of crossover codes do not depend on any aspect about the sequence being encoded. This distinguishes crossover, for instance, from representations that grow more or less crowded with the complexity of the sequence they represent.

\subsection{Proof: Sequence information is i.i.d. across components}

Since the components for one sequence $Y$ are i.i.d. (\ref{sec:prop-comp-iid}) Therefore, while the $x^Y_j$ and $x^{Y'}_j$ might be dependent (e.g. if $Y$ and $Y'$ are very similar; see below), there can be no dependencies between $x^Y_j$ and $x^{Y'}_{j'}$ for $j \neq j'$. Therefore $\{x^{Y_1}_j, ..., x^{Y_{N_Y}}_j\}$ is independent across $j$. Since $I_j$ is a function of $\{x^{Y_1}_j, ..., x^{Y_{N_Y}}_j\}$, $I_j$ is independent across $j$ also. Since no $j$ is treated specially in our algorithm, $I_j$ must also be identically distributed, so given $P(Y)$, $I_j$ is i.i.d.

\subsection{Code similarity derivations}

Since $x^Y_j$ are i.i.d. across $j$ (\ref{sec:prop-comp-iid}) it follows that $s(X^Y, X^{Y'}) \sim \frac{1}{N}\textrm{Bi}(N, q)$, where $\textrm{Bi}(N, q)$ is a binomial distribution with success probability $q = P(x^Y_j = x^{Y'}_j)$. Thus, $\textrm{E}[s(X^Y, X^{Y'})] = q$ and $\textrm{Var}[s(X^Y, X^{Y'})] = q(1-q)/N$, and it suffices to only compute $q$. For conciseness, we drop the $j$ and write $x \equiv x_j$ from now on.

We define and compute a few useful quantities that will keep our calculations more precise.

@@

\subsubsection{Derivation: Similarity between symbol and sequence codes}

$$P(x^Y = x^i) = P(x^Y \leftarrow i)P(x^Y = x^i|x^Y \leftarrow i) + P(x^Y \not \leftarrow i)P(x^Y = x^i|x^Y \not \leftarrow i)
= P(x^Y \leftarrow i) + P(x^Y \not \leftarrow i)\frac{1}{Z}$$

@@

\subsubsection{Derivation: Similarity between codes for two sequences}

Let $u^i$ index the times $i$ appears in $Y_{1:t}$, $v^i$ the times $i$ appears in $Y_{t+1:L}$, and $(v')^i$ the times $i$ appears in $Y'_{t+1:L'}$. Let $n^i_u$ count how many times $i$ appears in $Y_{1:t}$, $n^i_v$ the number of times $i$ appears in $Y_{t+1:L}$, and $n^i_{v'}$ the number of times $i$ appears in $Y'_{t+1:L'}$. We also write these in vector format as $\mathbf{n}_u$, $\mathbf{n}_v$, and $\mathbf{n}_{v'}$, respectively.

@@

The event $x^Y = x^{Y'}$ can occur in 4 distinct ways, whose probabilities sum to $P(x^Y = x^{Y'})$. At present we assume a negligible chance two sequence or symbol codes are exactly identical.

@@

$$= [1 - (t+1)^\gamma f^{\gamma,L}_t][1 - (t+1)^\gamma f^{\gamma,L'}_t]P(x^Y = x^{Y'}|x^Y, x^{Y'} \not\leftarrow Y_{1:t})$$

@@

\textbf{Sum over cases}

@@

When $\gamma = 1$ this simplifies to

$$P(x^Y = x^{Y'}) =  \frac{t+1}{L+1}\frac{t+1}{L'+1}$$

$$
+ \frac{L-t}{L+1}\frac{L'-t}{L'+1}\left[\frac{\mathbf{n}^T_v\mathbf{n}_{v'}}{(L-t)(L'-t)} + \left(1 - \frac{\mathbf{n}^T_v\mathbf{n}_{v'}}{(L-t)(L'-t)} \right)\frac{1}{Z}\right]
$$

$$+\frac{t+1}{L+1}\frac{L'-t}{L'+1}\left[
\frac{\mathbf{n}^T_{u}\mathbf{n}_{v'}}{(t+1)(L'-t)}+ \left(1 - \frac{\mathbf{n}^T_{u}\mathbf{n}_{v'}}{(t+1)(L'-t)} \right)\frac{1}{Z}
\right]$$
$$
+\frac{L'-t}{L+1}\frac{t+1}{L'+1}\left[
\frac{\mathbf{n}^T_{v}\mathbf{n}_{u}}{(L-t)(t+1)} + \left(1 - \frac{\mathbf{n}^T_{v}\mathbf{n}_{u}}{(L-t)(t+1)} \right)\frac{1}{Z}
\right].
$$
